# Cross-references {#cross}

Cross-references make it easier for your readers to find and link to elements in your book.

## Chapters and sub-chapters

```{r}
library(vegnasis)
veg.raw <- vegnasis::nasis.veg
veg <- clean.veg(veg.raw)
veg <- fill.hts.df(veg)
x.cross <- summary.crown.thickness(veg, breaks = c(0:30))
nplots <- x.cross$plot |> unique() |> length()
x.cross2 <- x.cross |> group_by(stratum, type) |> mutate(cover = sum(Cover)/nplots)

```


There are two steps to cross-reference any heading:

1. Label the heading: `# Hello world {#nice-label}`. 
    - Leave the label off if you like the automated heading generated based on your heading title: for example, `# Hello world` = `# Hello world {#hello-world}`.
    - To label an un-numbered heading, use: `# Hello world {-#nice-label}` or `{# Hello world .unnumbered}`.

1. Next, reference the labeled heading anywhere in the text using `\@ref(nice-label)`; for example, please see Chapter \@ref(cross). 
    - If you prefer text as the link instead of a numbered reference use: [any text you want can go here](#cross).

## Captioned figures and tables

Figures and tables *with captions* can also be cross-referenced from elsewhere in your book using `\@ref(fig:chunk-label)` and `\@ref(tab:chunk-label)`, respectively.

See Figure \@ref(fig:nice-fig).

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center', fig.alt='Plot with connected points showing that vapor pressure of mercury increases exponentially as temperature increases.'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```




Don't miss Table \@ref(tab:nice-tab).



## Chapters and sub-2

There are two steps to cross-reference any heading:


```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(x.cross2, 10), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

More words

```{r}
library(ggplot2)

par(mar = c(4, 4, .1, .1))

ggplot(data=x.cross2, aes(x=type, y=stratum, size=cover, color=type))+
  geom_point()


```
