# Composition


This analysis using vegnasis package explores vegetation composition.

## Data Processing

```{r warning=FALSE, message = FALSE}
# remotes::install_github("phytoclast/vegnasis", dependencies = FALSE)
library(vegnasis)
library(ggplot2)

veg.raw <- vegnasis::nasis.veg
#Clean data
veg <- clean.veg(veg.raw)
#Select only Wexford County Michigan records, and exclude imprecise non bionomial (genus or family) records by requiring a blank space in the name.
veg <-  veg |> subset(grepl('MI165',plot) & grepl('[[:blank:]]',taxon))
#These tasks fill in missing plant 'types' and establishes the crown heights based on user inputs of stratum, live canopy heights, and taxon norms when user data is missing.
veg <- veg |> fill.type.df() |> fill.hts.df()

knitr::kable(
  head(veg, 10), booktabs = TRUE,
  caption = 'A table of the first 10 rows of the veg data.'
)
```
## Harmonizing Taxa

When combining data from multiple sources, the possibility exists that the ecologist used a different scientific name for a taxon. Before conducting cluster analysis, it is wise to harmonize the taxonomy. 

**harmonize.taxa()** 

- This function synonymizes taxa with [BONAP](http://bonap.org/) or for Mexican only plants, Kew's [Plants of the World Online](https://powo.science.kew.org/) circa 2022, and returns a binomial. Assuming that the ecologist has the correct concept for the plant while in the field, and is using a legitimate name for that concept, the name can be synonymized. This ensures that various vegetation summarizing functions work off a shared taxonomic backbone, and not misconstrue synonyms as plant diversity. 

## Associations

```{r warning=FALSE, message = FALSE}
#This function ranks each taxon by cover and retains the top 2-5 taxa, with fewer taxa retained with greater unequal dominance. Taxa are separated by '/' when different stratum, and '-' when in the same stratum.
veg.associations <- veg |> get.assoc()

knitr::kable(
  veg.associations, booktabs = TRUE,
  caption = 'A table with the top 2-5 taxa organized by stratum of each plot.'
)

```


## Cluster Analysis

```{r warning=FALSE, message = FALSE}
#Create plot matrix, based log transformed relative cover values.
m <- veg |> make.plot.matrix(tr = 'log', rc = TRUE)

knitr::kable(
  m[1:10,1:10], booktabs = TRUE,
  caption = 'A table of the first 10 rows of the community data matrix.'
)



#distance matrix based on Bray-Curtis simularity.
d = vegan::vegdist(m, method='bray')
#Cluster analysis using Ward's method using distance matrix.
t <- cluster::agnes(d, method = 'ward')|> as.hclust()
#Define number of groups to color the dendrogram by.
k = 3
groups <- cutree(t, k = k)
#This function rearranges the branchs and groups so that the tree is always oriented with most nested branches to the bottom of the plot (when tree oriented vertically with branches to the right).
groups <- dendrogrouporder(t, groups)
a = 'Vegetation of Wexford County'

```



```{r warning=FALSE, message = FALSE, fig.cap = "Dendrogram of selected plots", out.width = '80%', fig.asp = 2}

plot.dendro(a,d,t,groups)

```
