# Vegetation Plant Profile Diagrams

The reason vegnasis package was created in the first place is to generate pictorial drawings of vegetation structure using plant shapes. Currently, only the most basic shapes have been developed as a proof of concept: generic broadleaf, conifer, shrub, forb, graminoid, palm, and fern. Over time, it is expected that more templates would be drawn in [Inkscape](https://inkscape.org/) or similar vector drawing software, and converted to xy coordinates that can be plotted with the [ggplot2](https://ggplot2.tidyverse.org/) package. The graphical function simply uses the geom_polygon() element within ggplot to draw individual plants, with colors fading with distance to simulate depth.

Before plotting any veg plot data, it first needs to be standardized with vegnasis::clean.veg() or vegnasis::pre.fill.veg(). It is highly recommended to limit data to a single plot before processing with vegnasis::grow_plants(), which will create the elements needed to draw the plants with the vegnasis::veg_profile_plot() function. 

## Data Processing

```{r warning=FALSE, message = FALSE}
# remotes::install_github("phytoclast/vegnasis", dependencies = FALSE)
library(vegnasis)
library(ggplot2)

veg.raw <-  vegnasis::nasis.veg
veg <- clean.veg(veg.raw)
veg.select <- subset(veg,  grepl('2022MI165021.P',plot))
plants <- grow_plants(veg.select)


```

## Plot Plant Profiles


### Northern Michigan hardwood forest

```{r warning=FALSE, message = FALSE, fig.cap = "Structure of northern Michigan hardwood forest."}

veg.select <- subset(veg,  grepl('2022MI165002.P',plot))
plants <- grow_plants(veg.select)

veg_profile_plot(plants)

```

### Transformed Y axis

```{r warning=FALSE, message = FALSE, fig.cap = "Structure of a northern Michigan forest (transformed)."}

veg_profile_plot(plants, 'sqrt', 5)

```


### Northern Michigan mixed forest

```{r warning=FALSE, message = FALSE, fig.cap = "Structure of northern Michigan mixed forest."}

veg.select <- subset(veg,  grepl('2022MI165023.P',plot))
plants <- grow_plants(veg.select)

veg_profile_plot(plants, unit='m',  skycolor = 'white', fadecolor = 'lightgray', gridalpha = 0.1, groundcolor = 'darkgray')


```

### Northern Michigan pine forest

Override default tree colors and shapes.

```{r warning=FALSE, message = FALSE, fig.cap = "Structure of northern Michigan pine forest."}

veg.select <- subset(veg,  grepl('2022MI165021.P',plot))

taxon <- c('Acer rubrum', 'Pinus resinosa')
crfill <- c(NA,"#80991A")
stfill <- c('gray',"#B36666")
crshape <- c(NA,'conifer2')
override <- data.frame(taxon=taxon,stfill=stfill,crfill=crfill,crshape=crshape)
veg.select <- veg.select |> left_join(override)

plants <- grow_plants(veg.select)

veg_profile_plot(plants)

```



### Washington conifer forest

```{r warning=FALSE, message = FALSE, fig.cap = "Structure of a Washington conifer forest."}

veg.select <- subset(veg,  grepl('2021WA031024',plot))
plants <- grow_plants(veg.select)
veg_profile_plot(plants, unit='m',  skycolor = 'white', fadecolor = 'lightgray', gridalpha = 0.1, groundcolor = 'darkgray')


```



### Generic oak savanna

Many parameters can be adjusted such as making the plot longer or changing sky color. Add unknown deciduous shrub with silvery green leaves (e.i. *Eleagnus* without actually saying it) by specifying  habit code "S.BD" and a crown fill color (find a the rgb color hex code that looks right). 

```{r warning=FALSE, message = FALSE, fig.cap = "Structure of a generic oak savanna"}

#Make up savanna data

thiscolor = rgb(0.6,0.9,0.7)

plot <- c('plot1')
taxon <- c('Quercus macrocarpa','UNK','Pteridium', 'Festuca', 'Andropogon', 'Liatris')
cover <- c(20,5,10,60,10,5)
crown.max <- c(15,4,1,0.6,2,0.4)
crfill <- c(NA,thiscolor,NA,NA,NA,NA)
dbh <- c(60,NA,NA,NA,NA,NA)
habit <- c(NA,'S.BD',NA,NA,NA,NA)
mydata <- data.frame(plot=plot,taxon=taxon, cover=cover, habit=habit, crown.max = crown.max, dbh.max = dbh, crfill=crfill)

veg <- mydata |> pre.fill.veg()
plants <- grow_plants(veg, plength=100) #Grow more plants on a longer 100 m plot (default was 50 m).
veg_profile_plot(plants, unit='m',  skycolor = rgb(0.8,0.98,1), fadecolor = 'lightgray', gridalpha = 0.1, groundcolor = rgb(0.55,0.45,0.2), xlim=c(0,100))

```

