# Vegetation Plant Profile Diagrams

The reason vegnasis package was created in the first place is to generate pictorial drawings of vegetation structure using plant shapes. Currently, only the most basic shapes have been developed as a proof of concept: generic broadleaf, conifer, shrub, forb, graminoid, palm, and fern. Over time, it is expected that more templates would be drawn in [Inkscape](https://inkscape.org/) or similar vector drawing software, and converted to xy coordinates that can be plotted with the [ggplot2](https://ggplot2.tidyverse.org/) package. The graphical function simply uses the geom_polygon() element within ggplot to draw individual plants, with colors fading with distance to simulate depth.

Before plotting any veg plot data, it first needs to be standardized with vegnasis::clean.veg() or vegnasis::pre.fill.veg(). It is highly recommended to limit data to a single plot before processing with vegnasis::prepare_strata(), which will create the elements needed to draw the plants with the vegnasis::veg_profile_plot() function. 

## Data Processing

```{r warning=FALSE, message = FALSE}
# remotes::install_github("phytoclast/vegnasis", dependencies = FALSE)
library(vegnasis)
library(ggplot2)

veg.raw <-  vegnasis::nasis.veg
veg <- clean.veg(veg.raw)
veg.select <- subset(veg,  grepl('2022MI165021.P',plot))
plants <- prepare_strata(veg.select)


```

## Plot Plant Profiles

### Northern Michigan conifer forest

```{r warning=FALSE, message = FALSE, fig.cap = "Structure of northern Michigan conifer forest."}

veg.select <- subset(veg,  grepl('2022MI165021.P',plot))
plants <- prepare_strata(veg.select)

veg_profile_plot(plants)

```


### Northern Michigan hardwood forest

```{r warning=FALSE, message = FALSE, fig.cap = "Structure of northern Michigan hardwood forest."}

veg.select <- subset(veg,  grepl('2022MI165002.P',plot))
plants <- prepare_strata(veg.select)

veg_profile_plot(plants)

```

### Transformed Y axis

```{r warning=FALSE, message = FALSE, fig.cap = "Structure of a northern Michigan forest (transformed)."}

veg_profile_plot(plants, 'sqrt', 5)

```


### Northern Michigan mixed forest

```{r warning=FALSE, message = FALSE, fig.cap = "Structure of northern Michigan mixed forest."}

veg.select <- subset(veg,  grepl('2022MI165023.P',plot))
plants <- prepare_strata(veg.select)

veg_profile_plot(plants, unit='m',  skycolor = 'white', fadecolor = 'lightgray', gridalpha = 0.1, groundcolor = 'darkgray')


```

### Washington conifer forest

```{r warning=FALSE, message = FALSE, fig.cap = "Structure of a Washington conifer forest."}

veg.select <- subset(veg,  grepl('2021WA031024',plot))
plants <- prepare_strata(veg.select)
veg_profile_plot(plants, unit='m',  skycolor = 'white', fadecolor = 'lightgray', gridalpha = 0.1, groundcolor = 'darkgray')


```



### Generic oak savanna

Many parameters can be adjusted.

```{r warning=FALSE, message = FALSE, fig.cap = "Structure of a generic oak savanna"}

#Make up savanna data
obsite <- c('plot1')
obsspp <- c('Quercus macrocarpa','Pteridium', 'Festuca', 'Andropogon', 'Liatris')
abund <- c(20,10,60,10,5)
hts <- c(15,1,0.6,2,0.4)
dbh <- c(60,NA,NA,NA,NA)
mydata <- data.frame(obsite=obsite, obsspp=obsspp, abund=abund, hts=hts, dbh=dbh)

#Identify columns containing data corresponding to standard column names.
mydata <- mydata |> mutate(taxon=obsspp, cover=abund, plot=obsite, crown.max = hts, dbh.max = dbh)
veg <- mydata |> pre.fill.veg()
plants <- prepare_strata(veg)
veg_profile_plot(plants, unit='m',  skycolor = rgb(0.8,0.98,1), fadecolor = 'lightgray', gridalpha = 0.1, groundcolor = rgb(0.55,0.45,0.2))

```

