# Composition


This analysis using vegnasis package explores vegetation composition.

## Data Processing

```{r warning=FALSE, message = FALSE}
# remotes::install_github("phytoclast/vegnasis", dependencies = FALSE)
library(vegnasis)
library(ggplot2)

veg.raw <- vegnasis::nasis.veg
# With a connection to NASIS via soilDB...
# veg.raw <- soilDB::get_vegplot_species_from_NASIS_db()


#Clean data
veg <- clean.veg(veg.raw)
#Select only Wexford County Michigan records, and exclude imprecise non binomial (genus or family) records by requiring a blank space in the name.
veg <-  veg |> subset(grepl('MI165',plot) & grepl('[[:blank:]]',taxon))
#These tasks fill in missing plant 'types' and establishes the crown heights based on user inputs of stratum, live canopy heights, and taxon norms when user data is missing.
veg <- veg |> fill.type.df() |> fill.hts.df()

options(knitr.kable.NA = '...')
knitr::kable(
  head(veg, 10), row.names = FALSE, booktabs = TRUE,
  caption = 'A table of the first 10 rows of the veg data.'
) |> kableExtra::column_spec(5,italic=T) |> kableExtra::scroll_box(width = "100%")|>
  kableExtra::kable_classic(full_width = F, html_font = "Cambria", position = "left")
```
## Harmonizing Taxa

When combining data from multiple sources, the possibility exists that the ecologist used a different scientific name for a taxon. Before conducting cluster analysis, it is wise to harmonize the taxonomy.

**harmonize.taxa()**

- This function synonymizes taxa with [BONAP](http://bonap.org/) or for Mexican only plants, Kew's [Plants of the World Online](https://powo.science.kew.org/) circa 2022, and returns a binomial. Assuming that the ecologist has the correct concept for the plant while in the field, and is using a legitimate name for that concept, the name can be synonymized. This ensures that various vegetation summarizing functions work off a shared taxonomic backbone, and not misconstrue synonyms as plant diversity.

## Associations

```{r warning=FALSE, message = FALSE}
#This function ranks each taxon by cover and retains the top 2-5 taxa, with fewer taxa retained with greater unequal dominance. Taxa are separated by '/' when different stratum, and '-' when in the same stratum.
veg.associations <- veg |> get.assoc()

knitr::kable(
  veg.associations, booktabs = TRUE,
  caption = 'A table with the top 2-5 taxa organized by stratum of each plot.'
) |>
  kableExtra::kable_styling(full_width = F)

```


## Cluster Analysis

```{r warning=FALSE, message = FALSE}
#Create plot matrix, based log transformed relative cover values.
m <- veg |> make.plot.matrix(tr = 'log', rc = TRUE)

knitr::kable(
  m[1:10,1:10], booktabs = TRUE,
  caption = 'A table of the first 10 rows of the community data matrix.'
)



#distance matrix based on Bray-Curtis simularity.
d = vegan::vegdist(m, method='bray')
#Cluster analysis using Ward's method using distance matrix.
t <- cluster::agnes(d, method = 'ward')|> as.hclust()
#Define number of groups to color the dendrogram by.
k = 3
groups <- cutree(t, k = k)
#This function rearranges the branchs and groups so that the tree is always oriented with most nested branches to the bottom of the plot (when tree oriented vertically with branches to the right).
groups <- dendrogrouporder(t, groups)
a = 'Vegetation of Wexford County'

```



```{r warning=FALSE, message = FALSE, fig.cap = "Dendrogram of selected plots", out.width = '80%', fig.asp = 2}

plot.dendro(a,d,t,groups)

```




## Phylogenetically Weighted Cluster Analysis

Sometimes there are several unknown species in a plot, which may or may not be the same as a species in other plots. These unknown species are often recorded as genera (e.g. *Rubus*), or if more ambiguous, as family (e.g. Poaceae). Creating a similarity matrix which rates commonly held genera and families may provide some partial credit to link plots related by unknown species. Alternatively, when comparing vegetation from more geographically distant biotic realms with few overlapping species, there may yet be value in giving credit to shared genera and families in as much as some ecological traits are phylogenetically conserved.

```{r warning=FALSE, message = FALSE}

#Create alternative higher taxon datasets and combine them.
veg.genera <- veg |> mutate(taxon = link.taxonomy(taxon, taxrank=1))
veg.families <- veg |> mutate(taxon = link.taxonomy(taxon, taxrank=2))
veg.combined <- rbind(veg,veg.genera,veg.families)

#Create plot matrix, based log transformed relative cover values.
m <- veg.combined |> make.plot.matrix(tr = 'log', rc = TRUE)




#distance matrix based on Bray-Curtis simularity.
d = vegan::vegdist(m, method='bray')
#Cluster analysis using Ward's method using distance matrix.
t <- cluster::agnes(d, method = 'ward')|> as.hclust()
#Define number of groups to color the dendrogram by.
k = 3
groups <- cutree(t, k = k)
#This function rearranges the branchs and groups so that the tree is always oriented with most nested branches to the bottom of the plot (when tree oriented vertically with branches to the right).
groups <- dendrogrouporder(t, groups)
a = 'Vegetation of Wexford County'

```



```{r warning=FALSE, message = FALSE, fig.cap = "Phylogenetically Weighted Dendrogram of selected plots", out.width = '80%', fig.asp = 2}

plot.dendro(a,d,t,groups)

```


## Non-Metric Multidimensional Scaling (NMDS)

### Species Composition Data

While dendrograms display a hierarchical relationship among sites, in may obscure the fact that sites within clusters can sometimes be more similar to sites in adjacent clusters than to some sites within their own cluster. An ordination is an alternative way to show relationships in two or more dimensions. Every species in a data set represents a potential axis which to display a relationship, but displaying more than two or three quickly becomes confusing. Some species share the same niche and therefore provide redundant information. Ordinations such as Non-Metric Multidimensional Scaling (NMDS) is a way of reducing redundancy and combining the shared information among species into fewer axes.



```{r warning=FALSE, message = FALSE}

#Nonmetric Multidimensional Scaling (NMDS) ----
library(vegan)
ndim <- 2 #number of dimensions
nmds <- vegan::metaMDS(m, k=ndim)
set.seed(6190)


#join NMDS data components
spts.es <- data.frame(plot=rownames(m)) |> cbind(m)  
pt.df <- vegan::scores(nmds, display='sites') |> as_tibble(rownames='sites')  
sp.df <- vegan::scores(nmds, display='species') |> as_tibble(rownames='species')

#narrow list of species and assign PLANTS Symbols for compact display on plot.
spsums <- veg |> group_by(plot, taxon) |> summarise(cover = vegnasis::cover.agg(cover)) |> subset(cover > 1) |> group_by(taxon) |> summarise(nplot = length(cover), maxcover = max(cover)) |> subset(nplot >= 2 | maxcover >=25)
sp.df <- sp.df |> mutate(sp1 = stringr::str_replace_all(species,'\\.', ' '), sp2 = vegnasis::fill.usda.symbols(sp1)) |> subset(sp1 %in% spsums$taxon)
```

Ordinations require a separate step to discover groups among the continuum of points. In this case we will use kmeans clustering using the NMDS values as inputs.

```{r warning=FALSE, message = FALSE}
#cut into clusters
mtx <- scores(nmds, display='sites')
nclust <- 3
kc1 <- mtx |> kmeans(nclust) 
kc1 <- kc1$cluster
pt.df <- pt.df |> mutate(kc = as.factor(paste0('cluster',kc1))) 

#convex hulls
fromclusts <- unique(pt.df$kc)
for(i in 1:nclust){#i=1
  thisclust <-fromclusts[i]
  thishull <- pt.df |> subset(kc %in% thisclust)
  chull0 <- chull(thishull$NMDS1, thishull$NMDS2)
  thishull  <- thishull[chull0,]
  if(i==1){kchull=thishull}else{kchull=rbind(kchull,thishull)}
};rm(thisclust,thishull,chull0)


library(ggplot2)

gp <- ggplot() +
  geom_polygon(data=kchull, aes(x=NMDS1,y=NMDS2, color=kc), alpha=0, linewidth=1)+
  geom_point(data=pt.df, aes(x=NMDS1,y=NMDS2, color=kc), alpha=0.5, size=2)+
  geom_point(data=sp.df, aes(x=NMDS1,y=NMDS2), color='blue')+
  geom_text(data=sp.df, aes(label=sp2, x=NMDS1,y=NMDS2), vjust = 0, nudge_y = 0.02, nudge_x = 0.05, color='blue', size=2)+
  coord_fixed()

```


A NMDS plot consists of points representing the relationship among sites or vegetation plots with respect to their species compositions, and points representing the relationship among individual species emergent among these sites. Additional information representing trends in abiotic properties of these sites can be displayed as arrows projected onto the same species compositional space (not shown in this example).

```{r warning=FALSE, message = FALSE, fig.cap = "Non-metric Multidimensional Scaling plot with 3 Clusters"}

gp

```

Assignment points into groups or clusters is often unpredictable, and may need interpretation. 

```{r warning=FALSE, message = FALSE}

#asign cluster to associations
veg2 <- veg |> left_join(data.frame(plot = pt.df$sites, NMDS_cluster = pt.df$kc)) |>
  mutate(stratum = case_when(ht.max > 5 ~ 'tree', 
                             ht.max > 0.5 & type %in% c('tree','shrub/vine') ~ 'shrub',
                             TRUE ~ 'herb')) |> 
  group_by(NMDS_cluster, stratum, taxon) |> summarise(ht.max = weighted.mean(ht.max,cover), ht.min = weighted.mean(ht.min,cover), cover = cover.agg(cover), type=first(type)) |> mutate(plot=NMDS_cluster)


clust.ass <- get.assoc(veg2)

```

In this case, we can interpret the clusters from after summarizing their dominant species, which cooresponds to mesophytic (moisture loving), pyrophytic (fire loving), and hydrophytic (wetland) vegetation types.

```{r warning=FALSE, message = FALSE}


knitr::kable(
  clust.ass, booktabs = TRUE,
  caption = 'NMDS Cluster Assignment'
)

```

This relabels the clusters according to our interpretations of those clusters.

```{r warning=FALSE, message = FALSE}

kchull <- kchull |> mutate(community = case_when(kc %in% 'cluster1' ~ 'hydrophytic',
                                                 kc %in% 'cluster2' ~ 'pyrophytic',
                                                 kc %in% 'cluster3' ~ 'mesophytic'))
pt.df <- pt.df |> mutate(community = case_when(kc %in% 'cluster1' ~ 'hydrophytic',
                                               kc %in% 'cluster2' ~ 'pyrophytic',
                                               kc %in% 'cluster3' ~ 'mesophytic'))

gp <- ggplot() +
  geom_polygon(data=kchull, aes(x=NMDS1,y=NMDS2, color=community, fill=community), alpha=0.1, linewidth=1)+
  geom_point(data=pt.df, aes(x=NMDS1,y=NMDS2, color=community), alpha=0.5, size=2)+
  geom_point(data=sp.df, aes(x=NMDS1,y=NMDS2), color='black')+
  geom_text(data=sp.df, aes(label=sp2, x=NMDS1,y=NMDS2), vjust = 0, nudge_y = 0.02, nudge_x = 0.05, color='black', size=2)+
  scale_color_manual(breaks = c("mesophytic", "pyrophytic", "hydrophytic"), values = c("green","red","blue"))+
  scale_fill_manual(breaks = c("mesophytic", "pyrophytic", "hydrophytic"), values = c("green","red","blue"))+
  coord_fixed()

```



```{r warning=FALSE, message = FALSE, fig.cap = "Non-metric Multidimensional Scaling plot with clusters interpreted"}

gp

```


### Environmental Data

Another element of NMDS analysis is the ability to relate gradients in species composition to gradients in abiotic properties of the sites. For this we need pedon data (or else various other gridded data environmental data extractable from the observation coordinates). Using the demo data, we need to use tools from the **aqp** package to extract the site and soil taxonomy data and the soil horizon data. We can then summarize horizons and soil taxonomy according to criteria useful to this area.


```{r warning=FALSE, message = FALSE}
library(soilDB)
library(aqp)
# s <- soilDB::fetchNASIS(from = "pedons",SS=T) #use this to fetch your own pedons from NASIS
s <- vegnasis::pedons20251212
sites <- aqp::site(s)


#extract horizon data from pedon collection
h <- aqp::horizons(s)
h <- h |> mutate(h50 = ifelse(hzdepb > 50, 50, hzdepb) - ifelse(hzdept > 50, 50, hzdept),
                 h150 = ifelse(hzdepb > 150, 150, hzdepb) - ifelse(hzdept > 150, 150, hzdept),
                 pH = phfield,
                 Bh = ifelse(grepl('Bh', hzname),1, 0),
                 Bs = ifelse(grepl('Bs', hzname),1, 0),
                 rock = ifelse(!is.na(hzname) & grepl('Cr|R', hzname),hzdept, 250),
                 carb = ifelse(!is.na(effclass) & !effclass %in% 'none', hzdept, 250)
)
#convert texture class to particle composition where missing
x <- texcl_to_ssc(texcl = h$texcl, clay = h$clay)
h <- h |> mutate(frag = ifelse(is.na(total_frags_pct), total_frags_pct, 0),
                 sand = ifelse(is.na(sand) & !is.na(x$sand), x$sand, sand),
                 silt = ifelse(is.na(silt) & !is.na(x$silt), x$silt, silt),
                 clay = ifelse(is.na(clay) & !is.na(x$clay), x$clay, clay),
                 # sand = ifelse(is.na(sand), 0, sand),
                 # silt = ifelse(is.na(silt), 0, silt),
                 # clay = ifelse(is.na(clay), 0, clay),
                 om = case_when(!is.na(lieutex) ~ 100 ,
                                grepl('A',hzname) ~ 1,
                                TRUE ~ 0))
#fixed depth weighted means
soil <- h |> group_by(peiid) |> summarise(frag50 = weighted.mean(frag, h50, na.rm = T),
                                          sand50 = weighted.mean(sand, h50, na.rm = T),
                                          sand150 = weighted.mean(sand, h150, na.rm = T),
                                          clay150 = weighted.mean(clay, h150, na.rm = T),
                                          frag150 = weighted.mean(frag, h150, na.rm = T),
                                          pH50 = weighted.mean(pH, h50, na.rm = T),
                                          OM150 = weighted.mean(om, h150, na.rm = T),
                                          rockdepth = min(rock, na.rm = T),
                                          carbdepth = min(carb, na.rm = T),
                                          Bh = max(Bh, na.rm = T),
                                          Bs = max(Bs, na.rm = T)) |> mutate(sand50 = ifelse(is.na(sand50), 0,sand50),
                                                                             sand150 = ifelse(is.na(sand150), 0,sand150),
                                                                             clay150 = ifelse(is.na(clay150), 0,clay150))
#reduce number of columns
soil <- subset(sites, select=c(siteiid, peiid, site_id, longstddecimaldegrees, latstddecimaldegrees, elev_field, taxonname, taxclname, taxorder, taxsubgrp, drainagecl, ecositeid, ecositenm, flodfreqcl, slope_field)) |> left_join(soil, by=join_by(peiid==peiid))

#Soil sorting criteria ----
soil <- soil |> mutate(pH50 = ifelse(is.nan(pH50), NA, pH50),
                       chem = case_when(grepl('od',taxclname) | grepl('ult',taxclname) | grepl('dys',taxclname) ~ 'acidic',
                                        grepl('alfs',taxclname)|grepl('euic',taxclname)|grepl('oll',taxclname) ~ 'nonacid',
                                        !is.na(pH50) & pH50 < 5.5 ~ 'acidic',
                                        carbdepth > 100 & grepl('psam',taxclname) ~ 'acidic',
                                        TRUE ~ 'nonacid'),
                       tex = case_when(OM150 > 50 | grepl('ist',taxclname) ~ 'mucky',
                                       (sand50 > 70 & sand150 > 80) | sand50 > 80 ~ 'sandy',

                                       TRUE ~ 'loamy'),
                       moisture = case_when(drainagecl %in% c("very poorly","poorly") ~ 'wet',
                                            drainagecl %in% c("somewhat poorly","moderately well") ~ 'moist',
                                            TRUE ~ 'dry'),
                       other = case_when(!flodfreqcl %in% 'none' ~ 'flood',
                                         grepl('od',taxclname) | grepl('alfs',taxclname)|grepl('euic',taxclname)|
                                           grepl('oll',taxclname)| (!is.na(pH50) & pH50 >= 6) |
                                           carbdepth < 100 | sand150 < 70 ~ 'rich',
                                         TRUE ~ 'other'))
#turn qualitative to quanitative ----
soil <- soil |> mutate(acidic = ifelse(chem %in% 'acidic',1,0),
                       hydric = case_when(drainagecl %in% c("very poorly","poorly") ~ 1,
                                          TRUE ~ 0),
                       draindepth = case_when(drainagecl %in% "very poorly" ~ 0,
                                      drainagecl %in% "poorly" ~ 25,
                                      drainagecl %in% "somewhat poorly" ~ 50,
                                      drainagecl %in% "moderately well" ~ 100,
                                      TRUE ~ 200),
                       north = latstddecimaldegrees,
                       east = longstddecimaldegrees,
                       elev = elev_field)

```

The environmental data needs to be aligned with the species matrix so that the environmental fit to the NMDS can be run correctly. The NMDS analysis is then recreated along with the clustering steps as done above.


```{r warning=FALSE, message = FALSE}

#align site ids from vegetation
vegsites <- veg.raw[,c('siteiid', 'vegplotid')] |> unique()
soil <- soil |> left_join(vegsites)
soil <- soil |> subset(!is.na(vegplotid), select=c(vegplotid,frag50, sand50,sand150,clay150,frag150,pH50,OM150,rockdepth,carbdepth,Bh,Bs,acidic,hydric,draindepth,north,east,elev))
#combine soil/site data to species matrix
m <- veg |> make.plot.matrix(tr = 'log', rc = TRUE)
sppcols <- colnames(m)
soilcols <- colnames(soil)[-1]
m$vegplotid <- rownames(m)
m <- m |> left_join(soil)
m.spp <- m[,sppcols]
m.env <- m[,soilcols]

#Conduct NMDS analysis
ndim <- 2
set.seed(6190)
nmds <- vegan::metaMDS(m.spp, k=ndim) 

#Fit abiotic attributes to NMDS scores.
en <- vegan::envfit(nmds, m.env, na.rm = TRUE, choices=c(1:ndim))
en.df <- vegan::scores(en, display='vectors')|> as_tibble(rownames='vectors')

#join NMDS data components
spts.es <- data.frame(plot=rownames(m.spp)) |> cbind(m.spp)
pt.df <- vegan::scores(nmds, display='sites') |> as_tibble(rownames='sites')
sp.df <- vegan::scores(nmds, display='species') |> as_tibble(rownames='species')

#narrow list of species and assign PLANTS Symbols for compact display on plot.
spsums <- veg |> group_by(plot, taxon) |> summarise(cover = vegnasis::cover.agg(cover)) |> subset(cover > 1) |> group_by(taxon) |> summarise(nplot = length(cover), maxcover = max(cover)) |> subset(nplot >= 2 | maxcover >=25)
sp.df <- sp.df |> mutate(sp1 = stringr::str_replace_all(species,'\\.', ' '), sp2 = vegnasis::fill.usda.symbols(sp1)) |> subset(sp1 %in% spsums$taxon)

#cut into clusters
mtx <- scores(nmds, display='sites')
nclust <- 3
kc1 <- mtx |> kmeans(nclust)
kc1 <- kc1$cluster
pt.df <- pt.df |> mutate(kc = as.factor(paste0('cluster',kc1)))

#convex hulls
fromclusts <- unique(pt.df$kc)
for(i in 1:nclust){#i=1
  thisclust <-fromclusts[i]
  thishull <- pt.df |> subset(kc %in% thisclust)
  chull0 <- chull(thishull$NMDS1, thishull$NMDS2)
  thishull  <- thishull[chull0,]
  if(i==1){kchull=thishull}else{kchull=rbind(kchull,thishull)}
};rm(thisclust,thishull,chull0)


#asign cluster to associations
veg2 <- veg |> left_join(data.frame(plot = pt.df$sites, NMDS_cluster = pt.df$kc)) |>
  mutate(stratum = case_when(ht.max > 5 ~ 'tree',
                             ht.max > 0.5 & type %in% c('tree','shrub/vine') ~ 'shrub',
                             TRUE ~ 'herb')) |>
  group_by(NMDS_cluster, stratum, taxon) |> summarise(ht.max = weighted.mean(ht.max,cover), ht.min = weighted.mean(ht.min,cover), cover = cover.agg(cover), type=first(type)) |> mutate(plot=NMDS_cluster)


clust.ass <- get.assoc(veg2)
veg.ass <- get.assoc(veg) |> left_join(data.frame(plot = pt.df$sites, NMDS_cluster = pt.df$kc)) |> arrange(NMDS_cluster, association)


kchull <- kchull |> mutate(community = case_when(kc %in% 'cluster1' ~ 'hydrophytic',
                                                 kc %in% 'cluster2' ~ 'pyrophytic',
                                                 kc %in% 'cluster3' ~ 'mesophytic'))
pt.df <- pt.df |> mutate(community = case_when(kc %in% 'cluster1' ~ 'hydrophytic',
                                               kc %in% 'cluster2' ~ 'pyrophytic',
                                               kc %in% 'cluster3' ~ 'mesophytic'))


#Add abiotic coorelations as arrows to the species and site plot as arrows.
gp <- ggplot() +
  geom_polygon(data=kchull, aes(x=NMDS1,y=NMDS2, color=community, fill=community), alpha=0.1, linewidth=1)+
  geom_point(data=pt.df, aes(x=NMDS1,y=NMDS2, color=community), alpha=0.5, size=2)+
  geom_point(data=sp.df, aes(x=NMDS1,y=NMDS2), color='darkgray')+
  geom_text(data=sp.df, aes(label=sp2, x=NMDS1,y=NMDS2), vjust = 0, nudge_y = 0.02, nudge_x = 0.05, color='darkgray', size=2)+
  geom_segment(data=en.df, aes(x=0,y=0,xend=NMDS1,yend=NMDS2), arrow = arrow(length = unit(0.03, "npc")), color='black')+
  geom_text(data=en.df, aes(label=vectors, x=NMDS1,y=NMDS2), vjust = 0, nudge_y = 0.02, nudge_x = 0.05, color='black')+
  scale_color_manual(breaks = c("mesophytic", "pyrophytic", "hydrophytic"), values = c("green","red","blue"))+
  scale_fill_manual(breaks = c("mesophytic", "pyrophytic", "hydrophytic"), values = c("green","red","blue"))+
  coord_fixed()

```



```{r warning=FALSE, message = FALSE, fig.cap = "Non-metric Multidimensional Scaling Environmental Biplot"}

gp

```
