# Composition


This analysis using vegnasis package explores vegetation composition.

## Data Processing

```{r warning=FALSE, message = FALSE}
# remotes::install_github("phytoclast/vegnasis", dependencies = FALSE)
library(vegnasis)
library(ggplot2)

veg.raw <- vegnasis::nasis.veg
# With a connection to NASIS via soilDB...
# veg.raw <- soilDB::get_vegplot_species_from_NASIS_db()


#Clean data
veg <- clean.veg(veg.raw)
#Select only Wexford County Michigan records, and exclude imprecise non binomial (genus or family) records by requiring a blank space in the name.
veg <-  veg |> subset(grepl('MI165',plot) & grepl('[[:blank:]]',taxon))
#These tasks fill in missing plant 'types' and establishes the crown heights based on user inputs of stratum, live canopy heights, and taxon norms when user data is missing.
veg <- veg |> fill.type.df() |> fill.hts.df()

options(knitr.kable.NA = '...')
knitr::kable(
  head(veg, 10), row.names = FALSE, booktabs = TRUE,
  caption = 'A table of the first 10 rows of the veg data.'
) |> kableExtra::column_spec(5,italic=T) |> kableExtra::scroll_box(width = "100%")|>
  kableExtra::kable_classic(full_width = F, html_font = "Cambria", position = "left")
```
## Harmonizing Taxa

When combining data from multiple sources, the possibility exists that the ecologist used a different scientific name for a taxon. Before conducting cluster analysis, it is wise to harmonize the taxonomy.

**harmonize.taxa()**

- This function synonymizes taxa with [BONAP](http://bonap.org/) or for Mexican only plants, Kew's [Plants of the World Online](https://powo.science.kew.org/) circa 2022, and returns a binomial. Assuming that the ecologist has the correct concept for the plant while in the field, and is using a legitimate name for that concept, the name can be synonymized. This ensures that various vegetation summarizing functions work off a shared taxonomic backbone, and not misconstrue synonyms as plant diversity.

## Associations

```{r warning=FALSE, message = FALSE}
#This function ranks each taxon by cover and retains the top 2-5 taxa, with fewer taxa retained with greater unequal dominance. Taxa are separated by '/' when different stratum, and '-' when in the same stratum.
veg.associations <- veg |> get.assoc()

knitr::kable(
  veg.associations, booktabs = TRUE,
  caption = 'A table with the top 2-5 taxa organized by stratum of each plot.'
) |>
  kableExtra::kable_styling(full_width = F)

```


## Cluster Analysis

```{r warning=FALSE, message = FALSE}
#Create plot matrix, based log transformed relative cover values.
m <- veg |> make.plot.matrix(tr = 'log', rc = TRUE)

knitr::kable(
  m[1:10,1:10], booktabs = TRUE,
  caption = 'A table of the first 10 rows of the community data matrix.'
)



#distance matrix based on Bray-Curtis simularity.
d = vegan::vegdist(m, method='bray')
#Cluster analysis using Ward's method using distance matrix.
t <- cluster::agnes(d, method = 'ward')|> as.hclust()
#Define number of groups to color the dendrogram by.
k = 3
groups <- cutree(t, k = k)
#This function rearranges the branchs and groups so that the tree is always oriented with most nested branches to the bottom of the plot (when tree oriented vertically with branches to the right).
groups <- dendrogrouporder(t, groups)
a = 'Vegetation of Wexford County'

```



```{r warning=FALSE, message = FALSE, fig.cap = "Dendrogram of selected plots", out.width = '80%', fig.asp = 2}

plot.dendro(a,d,t,groups)

```




## Phylogenetically Weighted Cluster Analysis

Sometimes there are several unknown species in a plot, which may or may not be the same as a species in other plots. These unknown species are often recorded as genera (e.g. *Rubus*), or if more ambiguous, as family (e.g. Poaceae). Creating a similarity matrix which rates commonly held genera and families may provide some partial credit to link plots related by unknown species. Alternatively, when comparing vegetation from more geographically distant biotic realms with few overlapping species, there may yet be value in giving credit to shared genera and families in as much as some ecological traits are phylogenetically conserved.

```{r warning=FALSE, message = FALSE}

#Create alternative higher taxon datasets and combine them.
veg.genera <- veg |> mutate(taxon = link.taxonomy(taxon, taxrank=1))
veg.families <- veg |> mutate(taxon = link.taxonomy(taxon, taxrank=2))
veg.combined <- rbind(veg,veg.genera,veg.families)

#Create plot matrix, based log transformed relative cover values.
m <- veg.combined |> make.plot.matrix(tr = 'log', rc = TRUE)




#distance matrix based on Bray-Curtis simularity.
d = vegan::vegdist(m, method='bray')
#Cluster analysis using Ward's method using distance matrix.
t <- cluster::agnes(d, method = 'ward')|> as.hclust()
#Define number of groups to color the dendrogram by.
k = 3
groups <- cutree(t, k = k)
#This function rearranges the branchs and groups so that the tree is always oriented with most nested branches to the bottom of the plot (when tree oriented vertically with branches to the right).
groups <- dendrogrouporder(t, groups)
a = 'Vegetation of Wexford County'

```



```{r warning=FALSE, message = FALSE, fig.cap = "Phylogenetically Weighted Dendrogram of selected plots", out.width = '80%', fig.asp = 2}

plot.dendro(a,d,t,groups)

```


## Non-Metric Multidimensional Scaling (NMDS)

While dendrograms display a hierarchical relationship among sites, in may obscure the fact that sites in within clusters can sometimes be more similar to site s in adjacent clusters than to some sites within the same cluster. An ordination displaying these relationships in two or more dimensions. Every species in a data set represents a potential axis which to display a relationship, but displaying more than two or three quickly becomes confusing. Some species share the same niche and therefore provide redundant information. Ordinations such as Non-Metric Multidimensional Scaling (NMDS) is a way of reducing redundancy and combining the shared information among species into fewer axes.



```{r warning=FALSE, message = FALSE}

#Nonmetric Multidimensional Scaling (NMDS) ----
library(vegan)
ndim <- 2 #number of dimensions
nmds <- vegan::metaMDS(m, k=ndim)
set.seed(6190)


#join NMDS data components
spts.es <- data.frame(plot=rownames(m)) |> cbind(m)  
pt.df <- vegan::scores(nmds, display='sites') |> as_tibble(rownames='sites')  
sp.df <- vegan::scores(nmds, display='species') |> as_tibble(rownames='species')

#narrow list of species and assign PLANTS Symbols for compact display on plot.
spsums <- veg |> group_by(plot, taxon) |> summarise(cover = vegnasis::cover.agg(cover)) |> subset(cover > 1) |> group_by(taxon) |> summarise(nplot = length(cover), maxcover = max(cover)) |> subset(nplot >= 2 | maxcover >=25)
sp.df <- sp.df |> mutate(sp1 = stringr::str_replace_all(species,'\\.', ' '), sp2 = vegnasis::fill.usda.symbols(sp1)) |> subset(sp1 %in% spsums$taxon)

#cut into clusters
mtx <- scores(nmds, display='sites')
nclust <- 3
kc1 <- mtx |> kmeans(nclust) 
kc1 <- kc1$cluster
pt.df <- pt.df |> mutate(kc = as.factor(paste0('cluster',kc1))) 

#convex hulls
fromclusts <- unique(pt.df$kc)
for(i in 1:nclust){#i=1
  thisclust <-fromclusts[i]
  thishull <- pt.df |> subset(kc %in% thisclust)
  chull0 <- chull(thishull$NMDS1, thishull$NMDS2)
  thishull  <- thishull[chull0,]
  if(i==1){kchull=thishull}else{kchull=rbind(kchull,thishull)}
};rm(thisclust,thishull,chull0)


library(ggplot2)

gp <- ggplot() +
  geom_polygon(data=kchull, aes(x=NMDS1,y=NMDS2, color=kc), alpha=0, linewidth=1)+
  geom_point(data=pt.df, aes(x=NMDS1,y=NMDS2, color=kc), alpha=0.5, size=2)+
  geom_point(data=sp.df, aes(x=NMDS1,y=NMDS2), color='blue')+
  geom_text(data=sp.df, aes(label=sp2, x=NMDS1,y=NMDS2), vjust = 0, nudge_y = 0.02, nudge_x = 0.05, color='blue', size=2)+
  coord_fixed()

```


A NMDS plot consists of points representing the relationship among sites or vegetation plots with respect to their species compositions, and points representing the relationship among individual species emergent among these sites. Additional information representing trends in abiotic properties of these sites can be displayed as arrows projected onto the same species compositional space (not shown in this example).

```{r warning=FALSE, message = FALSE, fig.cap = "Non-metric Multidimensional Scaling plot with 3 Clusters"}

gp

```

Assignment points into groups or clusters is often unpredictable, and may need interpretation. 

```{r warning=FALSE, message = FALSE}

#asign cluster to associations
veg2 <- veg |> left_join(data.frame(plot = pt.df$sites, NMDS_cluster = pt.df$kc)) |>
  mutate(stratum = case_when(ht.max > 5 ~ 'tree', 
                             ht.max > 0.5 & type %in% c('tree','shrub/vine') ~ 'shrub',
                             TRUE ~ 'herb')) |> 
  group_by(NMDS_cluster, stratum, taxon) |> summarise(ht.max = weighted.mean(ht.max,cover), ht.min = weighted.mean(ht.min,cover), cover = cover.agg(cover), type=first(type)) |> mutate(plot=NMDS_cluster)


clust.ass <- get.assoc(veg2)

```

In this case, we can interpret the clusters from a summarizing their dominant species which cooresponds to mesophytic (moisture loving), pyrophytic (fire loving), and hydrophytic (wetland) vegetation types.

```{r warning=FALSE, message = FALSE}


knitr::kable(
  clust.ass, booktabs = TRUE,
  caption = 'NMDS Cluster Assignment'
)

```

This relabels the clusters according to our interpretations of those clusters.

```{r warning=FALSE, message = FALSE}

kchull <- kchull |> mutate(community = case_when(kc %in% 'cluster1' ~ 'hydrophytic',
                                                 kc %in% 'cluster2' ~ 'pyrophytic',
                                                 kc %in% 'cluster3' ~ 'mesophytic'))
pt.df <- pt.df |> mutate(community = case_when(kc %in% 'cluster1' ~ 'hydrophytic',
                                               kc %in% 'cluster2' ~ 'pyrophytic',
                                               kc %in% 'cluster3' ~ 'mesophytic'))

gp <- ggplot() +
  geom_polygon(data=kchull, aes(x=NMDS1,y=NMDS2, color=community, fill=community), alpha=0.1, linewidth=1)+
  geom_point(data=pt.df, aes(x=NMDS1,y=NMDS2, color=community), alpha=0.5, size=2)+
  geom_point(data=sp.df, aes(x=NMDS1,y=NMDS2), color='black')+
  geom_text(data=sp.df, aes(label=sp2, x=NMDS1,y=NMDS2), vjust = 0, nudge_y = 0.02, nudge_x = 0.05, color='black', size=2)+
  scale_color_manual(breaks = c("mesophytic", "pyrophytic", "hydrophytic"), values = c("green","red","blue"))+
  scale_fill_manual(breaks = c("mesophytic", "pyrophytic", "hydrophytic"), values = c("green","red","blue"))+
  coord_fixed()

```



```{r warning=FALSE, message = FALSE, fig.cap = "Non-metric Multidimensional Scaling plot with clusters interpreted"}

gp

```

